---
title: "Boreogadus Saida (Arctic Cod) - Alaska/Canada"
author: "Natasha Howe"
date: "April 2025"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load libraries
library(tidyverse)

```

EXPLANATION OF PRIOR WORK

### OUT OF ORDER WORK

LD Pruning:

```{bash, ld_pruning, eval=F, echo=T}
sbatch scripts/boreogadus_peak_ld_pruning.sh -d /home/nhowe/arctic -k scripts/chr1-3_peak_input.txt -b boreogadus_wholegenome_20241222.beagle.gz -f /figures/ -p boreogadus -m boreogadus_filtered_downsampled_bamslist.txt -l no
```

# Project Template - Not Filled Out

## Introduction

EXPLANATION OF PROJECT

## Sample Collection

LIST POPULATIONS, MAKE TABLE?

```{r MetaData, eval=F, echo=T}

# All samples and sequencing information.
read.csv("./data/raw/METADATA.csv")
```

## Whole Genome Sequencing

EXPLANATION OF LAB WORK AND SEQUENCING EFFORTS

## Data Analysis Pipeline

Some resources came from the lcWGS analysis pipeline was developed by
Laura Timm (<https://github.com/letimm/WGSfqs-to-genolikelihoods>) based
on guides from Dr. Nina Therkildsen lab. The pipeline is undergoing
iterative updates on the AFSC ABL github repository
(<https://github.com/AFSC-Genetics/lcWGS-pipeline>) which is currently
not public, but will become so when there is a stable version of the
pipeline.

### Access to Fastq Files

WHERE DID DATA COME FROM

Uploaded to Matt's analysis computer.

```{bash SEDNA_Txfr, eval=F, echo=T}
#transfer

#MD5SUM after transfer

nohup md5sum -c seqmatic_fastq_md5.sum > NOA002_seqmatic_sedna.md5out
```

Then create a file with a list of all the fastq files and their paths,
which will be referenced in lcWGS_boreogadus_config.txt.

```{bash FqFilePaths, eval=F, echo = T}
cd ../../scratch2/nhowe/runtiming/boreogadus

ls -d $PWD/*_R*.fq.gz > boreogadus_fastqs.txt

# move to project folder
cp boreogadus_fastqs.txt ~/runtiming/boreogadus/scripts/
```

In this project, I had a folder for each population, which was
unnecessary. Therefore, I did this in each pop folder and then combined.

### Step 0: Setup

```         
IF THIS IS FIRST TIME USING THE PIPELINE:
SETUP VIRTUAL ENVIRONMENT AND DOWNLOAD MULTIQC VERSION 1.14!!
NEWER VERSIONS OF MULTIQC WONT RUN ON PYTHON VERSIONS < 3.8
```

```{bash PipInstall, eval=T, echo=T}
pip install multiqc==1.14
```

The first thing we need to do is download a reference genome and move it
over to sedna so that we can align our data. We are using the Arctic cod
(*boreogadus saida*) reference genome.

```         
GCA_964260565.1
```

With reference genome in hand (ncbi_dataset.zip) Unzip it locally to a
directory named ./Ref_genomes/boreogadus.

```{bash Unzip, eval=F, echo=T}
unzip ncbi_dataset.zip -d ./Ref_genomes/boreogadus
```

Then copied just the fna file to sedna

```{bash Trsfr_chr, eval=F, echo=T}
scp ./chromosomes.txt user.name@161.55.52.157:./runtiming/boreogadus
```

This puts the actual reference genome down a few directories. You may
have downloaded the RefSeq (GCA), GenBank (GCF), or both formats.
Dealing with the GCF file is slightly easier down the line if you want
to read the genome into memory.

We can make a list of chromosomes by using a few commands in the
terminal. All the chromosome lines start with NC, so we can use grep to
find lines that start with (\^) the characters '\>NC' and pass those
lines to awk to print the first column (\$1) and pass that to sed to
remove the character '\>' that begins each line. We should have a text
file with all the chromosome names.

```{bash Chromosomes, eval=F, echo=T}
grep '^>NC' ./home/nhowe/reference_genomes/boreogadus/GCF_023373465.1_Oket_V2_genomic.fna |  awk '{print $1}' | sed 's/>//' > chromosomes_all.txt
```

boreogadus have 37 chromosomes + 1 mtDNA. LRRC9 is on Chr 35.

Note that if you use the Genbank file you will have to replace NC with
CM in the grep argument. Sometimes the mitochondrial genome is not
present in the GCA file but is in the GCF file so you may have to omit
it from our list of chromosomes to analyze.

```{r RmMtDNA, eval=F, echo=T}
sed '/NC_017838.1/d' chromosomes_all.txt > chromosomes.txt
```

Let's take a quick look

```{bash ChromHead, eval=T, echo=T}
vim chromosomes.txt 
```

We second need to create a config file (lcWGS_boreogadus_config.txt)
that will be passed to all of the python scripts to create the slurm
control files (.sh files) that we will run with bash command. This tab
delim file has a single line with a few parameters (mostly file paths)
for the pipeline such as the fastqs, adapter sequence, chromosomes,
reference genome, working directory. It also has a place to define a run
prefix and an email for job updates from the SLURM manage. Do not
include an underscore in your prefix!!

```{bash ConfigFile, eval=T, echo = T}
vim lcWGS_boreogadus_config.txt
```

Now that all that is done, we can start running the python scripts that
are part of the pipeline. The first script checking for required inputs,
generates some directories to house scripts and output, creates a
checkpoint file (boreogadus.ckpt) that keeps track of where we are in
the pipeline.

So we can run the python script with

```{bash, Step0_Configure, eval=F, echo=T}
python3 lcWGSpipeline_step0-configure.py -c lcWGS_boreogadus_config.txt
```

which outputs two shell files bwa-index_script.sh and fai_script.sh.
These two files do not need to finish before proceeding to step 1 of the
pipeline. Get size of chromosomes

```{bash ChromHead, eval=T, echo=T}
 cut -f1,2 GCF_006149115.2_Oner_1.1_genomic.fna.fai | grep '^NC' > chrom_sizes.txt
```

```{bash, Step0_bwa_fai, eval=F, echo=T}
./GCF_006149115.2_Oner_1.1_genomic_bwa-index_script.sh
./GCF_006149115.2_Oner_1.1_genomic_fai_script.sh
```

### Step 1: Quality Control

This step is a quality control of the sequencing data (raw fastqs). It
runs FastQC [@Andrews2010] on the raw sequence data. FastQC is a java
based application that runs a set of quality control measures and

-   Import data from BAM, SAM or FastQ files (any variant)
-   Provides a quick overview indicating in which areas there may be
    problems
-   Summary graphs and tables to quickly assess data quality
-   Export of results to an HTML based permanent report
-   Offline operation to allow automated generation of reports without
    running the interactive application

The shell scripts can be created with the python script step1-fastqc
with the command

```{bash, Step1, eval=F, echo=T}
python3 lcWGSpipeline_step1-qc.py -p boreogadus.ckpt
```

passing the checkpoint file with the -p flag. The output of this script
is two shell scripts boreogadus-raw_fastqcARRAY.sh and
boreogadus-raw_multiqcSLURM.sh. These two scripts must be run in
succession.

```{bash, Step1_fastqc, eval=F, echo = T}
./boreogadus-raw_fastqcARRAY.sh
```

```{bash, Step1_multiqc, eval=F, echo = T}
./boreogadus-raw_multiqcSLURM.sh
```

### Step 2: Trim

Step 2 of the pipeline trims adapters from raw fastqs with the program
TRIMMOMATIC [@Bolger2014] and quality-checks the trimmed fastqs with
multiQC [@Ewels2016]. The following python script outputs three shell
scripts boreogadus_trimARRAY.sh, boreogadus-trim_fastqcARRAY.sh, and
boreogadus-trim_multiqcSLURM.sh which should all be run one after the
other.

```{bash, Step2_Trim, eval=F, echo=T}
python3 lcWGSpipeline_step2-trim.py -p boreogadus.ckpt
```

```{bash, Step2_trim, eval=F, echo = T}
./boreogadus_trimARRAY.sh
```

```{bash, Step2_trim_fastqc, eval=F, echo = T}
./boreogadus-trim_fastqcARRAY.sh
```

```{bash, Step2_trim_multiqc, eval=F, echo=T}
./boreogadus-trim_multiqcSLURM.sh
```

It is advisable to download the multiQC.html file and evaluate the data
quality before proceeding.

### Step 3: Align

Step 3 aligns reads to the reference genome that we downloaded with BWA
and runs the aligned reads through SAMTOOLS [@Li2009]: 'fixmate' cleans
up the read pairings and flags from BWA; a pair of 'view' statements
converts the .sam file to a .bam file and filters the .bam file for
non-unique and poor quality mappings; and 'sort' sorts the read pairings
by coordinate (instead of read name). After a .bam file is built,
duplicate reads are removed and (if the data is PE) overlapping reads
are clipped to generate the final .bam.

```{bash, Step3_Align, eval=F, echo=T}
python3 lcWGSpipeline_step3-align.py -p boreogadus.ckpt
```

```{bash, Step3_align, eval=F, echo=T}
./boreogadus_alignARRAY.sh
```

```{bash, Step3_depths, eval=F, echo=T}
./boreogadus_depthsARRAY.sh
```

There is an output file boreogadus_depths.csv which we can import and
run through the downsampling script to see if there is uneven
distribution of depths across early and late individuals. The depth of
these samples were quite low, and we didn't even use a cutoff because
the two individuals that were less than 0.5x were in the 0.4s.

```{R, plotDepths, eval=F, echo=T}
downsampling_boreogadus.R
```

##### Mapped Reads

Calculate the total number of mapped reads from the bamfiles with
samtools.

```{bash, mappedReads, eval=F, echo=T}
./boreogadus_mapped_reads.sh
```

Now calculate the average and standard deviation for the mapped reads.

```{r, mapped_reads, eval=F, echo=T}
mean(read.csv("./results/depth/boreogadus_mapped_reads.csv", header = F)$V1)
sd(read.csv("./results/depth/boreogadus_mapped_reads.csv", header = F)$V1)
```

### Step 4: Genotype Likelihoods

There is a batch effect in these data. Need to make sure that the
filters remove that batch effect.

Calculates genotype likelihoods for putatively polymorphic sites and all
sites. This is done on a chrom-by-chrom level using an array.

```{bash, Step4-GenoLikelihoods, eval=F, echo=T}
python3 lcWGSpipeline_step4-data.py -p boreogadus.ckpt -b boreogadus_blocklist_none.txt
```

Both of these scripts use boreogadus_angsdARRAY_input.txt (array file
with chromosomes) and a list of the bam files to analyze (non-blocklist
files -filtered_bamslist.txt).

```{bash, Step4_LikelihoodCalc, eval=F, echo=T}
# did not run globalArray
# cp output and create new output with specified filters tomorrow
cp boreogadus_polymorphicARRAY.sh boreogadus_minInd0.3_glsARRAY.sh
./boreogadus_minInd0.3_glsARRAY.sh
```

For calculating the genotype likelihoods we are making some implicit
decisions by running the canned pipeline script:

-   *minMapQ* [20]: the minimum map quality is 20. If a different
    mapping algorithm is used this value should be reassessed.
-   *C* [50]: use an adjustment of the Map Quality for excessive
    mismatches of 50. This is the suggested value for BWA and leads to
    fewer false positive variant calls
-   *minDepth* [48]: the minimum depth is set to half the number of
    individuals in the dataset. Discard site if total sequencing depth
    (all individuals added together) is below 0.5\*sample size.
-   *setMaxDepth* [960]: maximum depth is set to the number of
    individuals times 10. Discard site if total sequencing depth (all
    individuals added together) is above the number of individuals times
    10. 
-   *SNP_pval* [1e-10]: only work with sites with a p-value from the
    likelihood test of less than 1e-10.
-   *Gl* [1]: use SAMtools genotype likelihood calling algorithm. One
    assumption of the SAMtools algorithm is that it assumes that errors
    are not independent, but that once a first error occurs at a certain
    site in an individual, a second error is more likely to occur at the
    same site
-   *minMaf* [0.05]: only work with sites with a maf above 0.05
-   *minInd* [28]: Set to 30% of sample size to include a proportion of
    individuals.

### Step 5: Collate

```{bash, Step5-collate, eval=F, echo=T}
python3 lcWGSpipeline_step5-collate.py -p boreogadus.ckpt
```

Again, copy and edit new files, including the output name

```{bash Concatenate, eval=F, echo=T}
./boreogadus_concat_minInd0.3_minDepthHalf_mafs.sh
./boreogadus_concat_minInd0.3_minDepthHalf_beagles.sh
```

### Step 6: Population Analyses - FST

The groups here were 1) fall and 2) summer, where the first column is
the ABLG sample and the second tab-delimited column is the group. Lines
from the group file would read like so:

```         
ABLG1234  summer
ABLG1235  fall
```

```{R, group_file, eval=F, echo=T}
#To create this file: 
metadata_df <- read.csv("./data/metadata/yukon_boreogadus_metadata.csv")

group_df <- metadata_df %>%
  mutate(sampleID = sub("^", "ABLG", ABLG)) %>%
  select(sampleID, Runtime)

write.table(group_df, "./sedna_files/fall_summer_groups.txt", col.names = F, row.names = F, quote = F, sep = "\t")  
```

Move file to sedna scripts folder.

```{bash, PopAnaly_py, eval=F, echo=T}
python3 lcWGSpipeline_pop-analyses.py -p boreogadus.ckpt -g fall_summer_groups.txt
```

The python script creates a bamlist for each group, and the shell script
contains every combination of comparisons for FST. First, angsd calls
GLs for each subgroup, but it uses the sites flag from the file created
in Step 5, so the SNPs being called will be the same as for the
polymorphic files from Step 4. Therefore, some of the filtering steps
have been removed, such as: *SNP_pval* and *minMaf*.

*setminDepth* and set*maxDepth* are included, but adjusted to the sample
size of the group. Added minInd to 70%, and changed minMapQ to 20.

```{bash, fst, eval=F, echo=T}
# change name to better explain the Fst run
mv boreogadus_popARRAY.sh boreogadus_fst_fall-summer_ARRAY.sh

./boreogadus_fst_fall-summer_ARRAY.sh
```

This resulted in only 187,667 SNPs across the genome. Rerunning without
minInd and lowering minMapQ back to 15.

```{bash, fst_redo, eval=F, echo=T}
./boreogadus_fst_fall-summer_noMinInd_ARRAY.sh
```

###### FINAL CHOSEN RUN:

*setminDepth* and set*maxDepth* are included, but adjusted to the sample
size of the group. Added minInd to 30%, minDepth to half, maxDepth to
10\* and minMapQ to 20.

```{bash, fst_redo, eval=F, echo=T}
./boreogadus_minInd0.3_minDepthHalf_fstARRAY.sh
```

Plot Whole genome FST and for each chromosome.

```{r, WG_FST_plots, eval=F, echo=T}
fst_manhattan_boreogadus.R
```

Plot local FST for two main regions of interest with genes.

```{r, localFST_plots, eval=F, echo=T}
fst_manhattan_peaks_with_genes.R
```

Honed in on lrrc9 region (chr35) and new estrogen receptor beta region
(chr 29).

### Step 7: Local PCAs

Create local PCAs in lrrc9 and esrb.

```{bash LocalpcAngsd, eval=F, echo=T}
lrrc9_pca_minInd0.3.sh
esrb_expand_pca_minInd0.3.sh # expand because not exact gene region
```

Plot pcas for two main regions of interest.

```{r, PCA_plots, eval=F, echo=T}
pca_esrb_boreogadus.R
pca_lrrc9_boreogadus.R
```

The output of these pca groups created putative haplotype groups for
these individuals as homozygous early, heterozygous, and homozygous
late. Homozygous individuals were collected into a textfile to be rerun
in sedna for an allelic FST.
