---
title: "boreogadus-selection-scan"
author: "Natasha Howe"
date: "`r Sys.Date()`"
output: html_document
---

PCANGSD Selection Scan

PCAngsd has two implemented selection scans based on (i) FASTPCA and (ii) pcadapt. The associated paper compares the methods with high and low coverage data ([Meisner et al 2021](https://pmc.ncbi.nlm.nih.gov/articles/PMC8480091/#Sec2)).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-metadata, warning=F, message=F}
library(stringr)
library(tools)
library(tidyverse)
library(cowplot)

chr_table <- read_tsv("../../data/R/boreogadus_chromosomes.txt", 
                      col_names = T, show_col_types = F) %>%
  rename(lg = chrName)
```

## Neutral Data

```{r naming}
path= "../../results/pca_selection/"
filename="boreogadus_unfused_neutral_alpha0.05"
outdir="../../figures/pca_selection/"
nEig=4
```

PCA

```{r}
color_df <- read.delim("../../data/R/color_metadata_allpops_flip.txt", header = T)
meta_df <- read.delim("../../data/R/boreogadus_sampleID_popID.txt", sep="\t", 
                      header = F, col.names = c("sampleID","Pop"))
bam_df <- read.table("../../data/bam/boreogadus_filtered_downsampled_bams.txt", header = F)

# read in the covariance matrix
cov <- read.table(paste0(path,filename,"_e",nEig,".cov"))
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)
  e_per <- e$values/sum(e$values) # percent explained by each component
  # varPC1 <- format(round((e$values[1]/sum(e$values)*100), 2), nsmall = 2) 
  # varPC2 <- format(round((e$values[2]/sum(e$values)*100), 2), nsmall = 2) 
  # varPC3 <- format(round((e$values[3]/sum(e$values)*100), 2), nsmall = 2)
  # varPC4 <- format(round((e$values[4]/sum(e$values)*100), 2), nsmall = 2) 
  
# Get ID and pop info for each individual
  bam_df <- bam_df %>%
    mutate(sampleID = basename(file_path_sans_ext(V1))%>% 
             gsub("_downsampled|boreogadus_|_sorted", "", .)) %>%
    left_join(meta_df, by = "sampleID") %>% 
    mutate(Pop = ifelse(is.na(Pop),"Iceland",Pop)) %>%
    dplyr::select(sampleID, Pop)
    
#combine row names (Pop info) with the covariance matrix
pca.vectors = as_tibble(cbind(bam_df, e_vectors))
```

### PCA

```{r}
# create mypalette
mypalette <- as.vector(color_df$Color2)
  names(mypalette) <- color_df$Pop # attach pop name to palette color

pca.vectors[,1:7] %>%
  pivot_longer(cols = starts_with("V"), names_to = "axis", values_to = "val") %>%
  group_by(sampleID) %>% # Replace with your ID column
  mutate(axis_next = lead(axis),
         val_next = lead(val),
         pair = paste(axis, "vs", axis_next)) %>%
  filter(!is.na(val_next)) %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
  # plot
  ggplot(aes(x = val, y = val_next)) +
  geom_point(mapping=aes(color = Pop),alpha = 0.6) +
  facet_wrap(~pair, scales = "free") +
  scale_color_manual(values = mypalette) +
  labs(x = "Lower PC Axis", y = "Higher PC Axis",
       title=filename) +
  theme_bw()

# ggsave(paste0(outdir,filename,"_PCs1-4.jpg"), height = 6, width = 8)
rm(pca.vectors, cov, e, e_per, e_vectors,mypalette)
```

### Weights

```{r}
# PC weights
weight_tmp <- read_delim(paste0(path,filename,"_e",nEig,".weights"), delim=" ",
                         col_names = c("pc1","pc2", "pc3","pc4"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(path,filename,"_e",nEig,".sites"), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- read_tsv(str_c(path,filename,".snp_list"), 
                     col_names = c("lg","pos"),
                     show_col_types = F) %>%
  slice(-1) # bc pcangsd removes the first row from the cut beagle file

# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") %>%
  left_join(chr_table, by="lg")

loadings_plot <- weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight), color=as_factor(chr)), size=0.1) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(0, 100, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
  theme_cowplot() +
  labs(x = "Position (Mb)",
       title = "Boreogadus: Neutral Dataset - PCAngsd Loadings") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,filename,"_PC_loadings.jpg"), 
       plot = loadings_plot,
       height = 16, width = 20)

rm(weight_tmp, weight_table)
```

Definitely seems to be inversions driving most of the pc weights form pc1, pc3, and pc4.

### Selection outliers

```{r}
############### nicolas lou rmd
selection_tmp <- read_delim(paste0(path,filename,"_e",nEig,".selection"), delim=" ",
                         col_names = F, show_col_types = F) %>% 
  as_tibble()

# combine selection output with sites
selection_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(selection_tmp) %>%
  pivot_longer(cols = 3:(ncol(selection_tmp)+2), names_to = "pc", values_to = "d") %>%
  mutate(p=pchisq(d, 1, lower.tail=FALSE)) %>%
  left_join(chr_table, by="lg") %>%
  group_by(pc) %>%
  mutate(bonferroni=p.adjust(p, method="bonferroni"),
         fdr=p.adjust(p, method="fdr")) %>%
  ungroup()

# Plot for PC1 and PC2
selection_plot <- selection_table %>%
  ggplot(aes(x=as.numeric(pos)/10^6, y=-log10(p), color=as_factor(chr))) +
  geom_point(size=0.1) +
  geom_point(data=selection_table %>% filter(fdr<0.01), color="red", size=0.2) +
  # geom_point(data=selection_table %>% slice_min(p, n=round(nrow(selection_tmp)/10000)),
  #            color="red", size=0.2) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(10, 90, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
    labs(x = "Position (Mb)",
       title = "Boreogadus: Neutral Dataset Selection Scan") +
  theme_cowplot() +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,filename,"_selection_scan.jpg"), 
       plot = selection_plot, height = 15, width = 20)

rm(selection_tmp)
```

Awesome! PC2 does not have outliers... which that looks like the most neutral trend from before anyways..

### Remove PCA Outlier SNP Regions from Neutral Dataset

It looks like outliers are generally concentrated on one region when found on a chromosome. This means I can remove loci that are within the range of the first and last outlier positions.

```{r}
outlier_regions <- selection_table %>% 
  filter(fdr<0.01) %>%
  group_by(pc,chr) %>%
  summarize(minPos = min(pos), maxPos=max(pos)) %>%
  ungroup() %>%
  arrange(chr, minPos)
  
filtered_sites <- snp_list %>%
  left_join(chr_table, by="lg") %>% 
  left_join(., outlier_regions, by = "chr", relationship = "many-to-many") %>%
  group_by(chr, pos) %>%
  mutate(within_range = any(pos >= minPos & pos <= maxPos, na.rm = TRUE)) %>%
  filter(!within_range) %>%
  ungroup() %>%
  dplyr::select(chr,lg,pos) %>%
  distinct()

print(paste(nrow(filtered_sites),"SNPs left.",
  nrow(snp_list)-nrow(filtered_sites),"SNPs were removed."))

filtered_sites %>%
  mutate(chrpos=str_c(lg,"_",pos)) %>%
  dplyr::select(chrpos) %>%
  write_tsv(str_c("../../data/R/",filename,"_e",nEig,"_rmPCAngsdOutliers.tsv"), col_names=F)
```

### Remove MDS Outlier Regions from Neutral Dataset

Matt shared his MDS outliers output with me. Import here.

```{r}
library(readxl)

#mds_df <- read_xlsx("../../results/mds/MDS_analysis.xlsx")
mds_sig_df <- read_xlsx("../../results/mds/MDS_windows_significant.xlsx") %>%
  mutate(minPos = Start,
         maxPos = Stop)

mds_filtered_sites <- filtered_sites %>%
  left_join(chr_table, by="lg") %>% 
  left_join(., outlier_regions, by = "chr", relationship = "many-to-many") %>%
  group_by(chr, pos) %>%
  mutate(within_range = any(pos >= minPos & pos <= maxPos, na.rm = TRUE)) %>%
  filter(!within_range) %>%
  ungroup() %>%
  dplyr::select(chr,lg,pos) %>%
  distinct()

print(paste(nrow(mds_filtered_sites),"SNPs left.",
  nrow(filtered_sites)-nrow(mds_filtered_sites),"SNPs were removed."))












```
