---
title: "boreogadus-mds-outliers"
author: "Natasha Howe"
date: "`r Sys.Date()`"
output: html_document
---

## Inversions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-metadata, warning=F, message=F}
library(stringr)
library(tools)
library(tidyverse)
library(cowplot)

chr_table <- read_tsv("../../data/R/boreogadus_chromosomes.txt", 
                      col_names = T, show_col_types = F) %>%
  rename(lg = chrName)
```

## Input Data

```{r naming}
path= "../../results/pca_selection/"
filename="boreogadus_unfused_neutral_alpha0.05"
outdir="../../figures/pca_selection/"
```

```{r}
meta_df <- read.csv("../../data/R/boreogadus_metadata.csv", header = T) %>%
  mutate(Population = ifelse(Population=='Southhampton Island', 
                             'Southampton Island', Population),
         Pop = case_when(Population == 'SE Baffin' ~ 'SEBaffin',
                         TRUE ~ gsub(' .*$','',Population))) %>%
  filter(Pop != 'Unknown')

color_df <- read.delim2("../../data/R/color_metadata_allpops_flip.txt", 
                        header = T, row.names = NULL, sep = "\t") %>%
  mutate(Pop = gsub('hh','h',Pop)) %>%
  left_join(meta_df %>% distinct(Population, Pop), by  = "Pop")

bam_df <- read.table("../../data/bam/boreogadus_filtered_downsampled_bams.txt", header = F)

mypalette <- as.vector(color_df$Color2)
  names(mypalette) <- color_df$Pop # attach pop name to palette color

# Get ID and pop info for each individual
bam_df <- bam_df %>%
  mutate(sampleID = basename(file_path_sans_ext(V1))%>% 
           gsub("boreogadus_|_downsampled|_sorted", "", .)) %>%
  left_join(meta_df, by = "sampleID") %>% 
  mutate(Pop = ifelse(is.na(Pop),"Iceland",Pop)) %>%
  dplyr::select(sampleID, Pop)
```

### MDS Outliers

Matt shared his MDS outliers output with me. Import here.

```{r}
library(readxl)
library(RColorBrewer)

mds_df <- read_xlsx("../../results/mds/MDS_analysis.xlsx")

offsets <- mds_df %>%
  mutate(CHROM=as.numeric(CHROM)) %>%
  group_by(CHROM) %>%
  dplyr::summarise(max_bp = max(POS, na.rm = T), 
                   .groups = "drop") %>%
  arrange(CHROM) %>%
  mutate(cumulative_bp = cumsum(max_bp),
         offset = lag(cumulative_bp, default = 0),
         chrMid = (max_bp/2 + offset))

# Join back to the main df and calculate cumulative position
mds_df1 <- mds_df %>%
  left_join(offsets, by = "CHROM") %>%
  mutate(cumPos = POS + offset) %>%
  dplyr::select(CHROM,POS,cumPos,starts_with("MD")) %>%
  dplyr::relocate(cumPos, .before="MDS1")

mds_df_longer <- mds_df1 %>%
  pivot_longer(cols = starts_with("MD"),
               names_to = "MDS",
               values_to = "value")

mds_outliers <- mds_df_longer %>%
  group_by(MDS) %>%
  mutate(outlier = value %in% boxplot.stats(value)$out)

out <- boxplot.stats(mds_df1$MDS1)$out
out_ind <- which(mds_df1$MDS1 %in% c(out))
out_ind

outliers <- mds_df1[out_ind,]
View(outliers)

nb.colos <- nrow(offsets)
mycolors <- colorRampPalette(brewer.pal(8,"Set1"))(nb.colos)
names(mycolors) <- offsets$CHROM
```

MDS 1 only

```{r}
ggplot(mds_df1) + 
  # rectangles colored by chrom
  geom_rect(data=offsets, 
            aes(xmin=offset,xmax=cumulative_bp,
                ymin=min(mds_df1$MDS1)-0.01, ymax=max(mds_df1$MDS1)+0.01),
            fill=mycolors,alpha=0.25)+
  # all mds points
  geom_point(data=mds_df1,aes(x=cumPos,y=MDS1,color=as.factor(CHROM)),
             alpha=0.75,cex=0.5)+
  # outlier mds points
  geom_point(data=outliers,aes(x=cumPos,y=MDS1),color="black",cex=0.5)+
  scale_x_continuous(breaks=offsets$chrMid, # offsets$cumulative_bp,
                     labels=offsets$CHROM,
                     expand = c(0.002, 0.002))+
  scale_y_continuous(expand = c(0.001, 0.001))+
  scale_color_manual(values=mycolors)+
  ylab("MDS1")+
  xlab("Chromosome Number")+
  theme_bw() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.text.x = element_text(face="bold",size=6))
```

The outputs looks the same, except that the outlier cutoff is more stringent in his plot.

All MDSs faceted

```{r}
faceted_plot <- ggplot(mds_outliers) + 
  # rectangles colored by chrom
  geom_rect(data = offsets, 
            aes(xmin = offset, xmax = cumulative_bp,
                ymin = -Inf, ymax = Inf, # Use -Inf/Inf to fill the whole height
                fill = as.factor(CHROM)), 
            alpha = 0.25, inherit.aes = FALSE)+
  # all mds points
  geom_point(data=mds_outliers %>% filter(outlier==F),aes(x=cumPos,y=value,color=as.factor(CHROM)),
             alpha=0.75,cex=0.5)+
  # outlier mds points
  geom_point(data=mds_outliers %>% filter(outlier==T),aes(x=cumPos,y=value),color="black",cex=0.5)+
  scale_x_continuous(breaks=offsets$chrMid,
                     labels=offsets$CHROM,
                     expand = c(0.002, 0.002))+
  scale_y_continuous(expand = c(0.001, 0.001))+
  facet_grid(rows=vars(MDS), scales="free_y")+
  scale_color_manual(values=mycolors)+
  scale_fill_manual(values=mycolors)+
  theme_bw() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.text.x = element_text(face="bold",size=6))

ggsave("../../figures/mds/boreogadus_mds_plots.jpeg",
       plot=faceted_plot, width = 20, height = 25)
```

# Filter Neutral SNPs

#### Known/Suspected Inversions

Incorporated the MDS outliers Matt tested for inversions & the inversions identified in Hoff Supplemental Table 2 of the pop gen paper.

```{r}
# collected tables that matt created and from Hoff's pop paper
invers_hale <- read_xlsx("../../results/mds/inversion_regions_Hale_and_Hoff.xlsx",sheet = "Hale") %>%
  mutate(chr=as.numeric(chr),startPos = as.numeric(startPos),
         endPos = as.numeric(endPos))
invers_hoff <- read_xlsx("../../results/mds/inversion_regions_Hale_and_Hoff.xlsx",sheet = "Hoff") %>%
  mutate(chr=as.numeric(chr),startPos = as.numeric(startPos),
         endPos =as.numeric(endPos))

inversions <- bind_rows(invers_hale,invers_hoff)

```

### PCAs

Create peak file

```{r}
invers_hale %>%
  left_join(., chr_table,by="chr") %>%
  mutate(inputFormat=paste(row_number(),
                           lg,
                           format(startPos,scientific=F),
                           format(endPos,scientific=F),
                           sep=":")%>%
           gsub(" ","",.)) %>%
  select(inputFormat) %>%
  write_tsv("../../data/R/bsaida_inversions_input.txt", col_names = F)
```

```{bash, eval=F}
sbatch scripts/boreogadus_inversion_pca_only.sh -d /home/nhowe/arctic -k scripts/bsaida_inversions_input.txt -b boreogadus_wholegenome_20241222.beagle -f /figures/ -p bsaida -m boreogadus_filtered_downsampled_bams.txt -l no
```

After importing locally

```{r}
inv_cov_list <- list.files(path = "../../results/pca/inversion", 
                            pattern = "*.cov", full.names = T) %>%
  set_names(nm = basename(file_path_sans_ext(.)) %>% 
              sub("bsaida_","",.)) %>%
  map(~read.table(.)) 

# Assuming your list of dataframes is named 'data_list'
inv_pca_list <- lapply(inv_cov_list, function(cov) {
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)[,1:3] # only save PCs 1-3
  return( as_tibble(cbind(bam_df, e_vectors)) )
})

inv_pca_df <- bind_rows(inv_pca_list, .id = "inversion") %>%
  mutate(inversion2=inversion) %>%
  separate_wider_delim(inversion2, delim="_", 
                       names = c("chrName","start","end")) %>%
  left_join(chr_table, join_by("chrName"=="lg")) %>%
  mutate(startMb = sub("s","",start) %>% as.numeric(.)/1e6,
         endMb = sub("e","",end) %>% as.numeric(.)/1e6,
         strip_title = paste('chr',chr,':',startMb,'-',endMb,'Mb'),
         Pop = factor(Pop, levels = color_df$Pop),
         strip_title = factor(strip_title, 
                              levels = unique(inv_pca_df$strip_title)))

inv_pca_df %>%
  ggplot() +
  geom_point(aes(x = V1, y = V2, color = Pop),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free") +
  scale_color_manual(values=mypalette)+
  theme_bw()+
  theme(strip.text = element_text(size=8),
        panel.grid = element_blank())

ggsave("../../figures/test.jpeg", plot = test, width = 20, height = 8)
```

#### Remove two individuals consistently staying at 0,0 (missing data)
```{r}
missingdata_inds <- inv_pca_df %>%
  filter(strip_title==unique(inv_karyo$strip_title)[17],
         V1 > -0.02, V1 < 0.02) %>%
  pull(sampleID)

missingness_bam <- bam_df %>%
  mutate(filter = ifelse(sampleID %in% missingdata_inds,0,1)) 

missingness_bam %>%
  select(filter) %>%
  write_tsv("../../data/pca_filter/boreogadus_pca_rm_missingness_inds.txt", 
            col_names = F)
```

### New Filtered PCA
```{r}
inv_cov_list <- list.files(path = "../../results/inversion/pca", 
                            pattern = "*filter.cov", full.names = T) %>%
  set_names(nm = basename(file_path_sans_ext(.)) %>% 
              sub("bsaida_","",.) %>% sub("_filter","",.)) %>%
  map(~read.table(.)) 

# Assuming your list of dataframes is named 'data_list'
inv_pca_list <- lapply(inv_cov_list, function(cov) {
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)[,1:3] # only save PCs 1-3
  return( as_tibble(cbind(missingness_bam %>% 
                            filter(filter==1), 
                          e_vectors)) )
})

inv_pca_df <- bind_rows(inv_pca_list, .id = "inversion") %>%
  mutate(inversion2=inversion) %>%
  separate_wider_delim(inversion2, delim="_", 
                       names = c("chrName","start","end")) %>%
  left_join(chr_table, join_by("chrName"=="lg")) %>%
  mutate(startMb = sub("s","",start) %>% as.numeric(.)/1e6,
         endMb = sub("e","",end) %>% as.numeric(.)/1e6,
         strip_title = paste('chr',chr,':',startMb,'-',endMb,'Mb'),
         Pop = factor(Pop, levels = color_df$Pop),
         strip_title = factor(strip_title, 
                              levels = unique(inv_pca_df$strip_title)))

inv_pca_df %>%
  ggplot() +
  geom_point(aes(x = V1, y = V2, color = Pop),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free") +
  scale_color_manual(values=mypalette)+
  theme_bw()+
  theme(strip.text = element_text(size=8),
        panel.grid = element_blank())

# ggsave("../../figures/inversions/filter/inversion_pcas_by_pop.jpeg", width = 20, height = 8)
```
### Heterozygosity

Plink - create bed file for each region and then run heterozygosity

```{bash}
sbatch ~/arctic/scripts/boreogadus_plink_inv_het.sh
```

```{r}
inv_het_list <- list.files(path = "../../results/inversion/het", 
                            pattern = "*.het", full.names = T) %>%
  set_names(nm = basename(file_path_sans_ext(.)) %>% 
              sub("boreogadus_","",.)) %>%
  map(~read.table(., header=T) %>%
        mutate(het = (N.NM. - O.HOM.) / N.NM.) %>%
        bind_cols(bam_df))

inv_het_df <- bind_rows(inv_het_list,.id="inversion") %>%
  dplyr::select(inversion, sampleID, het) %>%
  drop_na(het) %>%
  left_join(inv_pca_df, ., by = c("inversion","sampleID")) %>%
  mutate(strip_title = factor(strip_title, 
                              levels = unique(inv_pca_df$strip_title))) %>%
  group_by(strip_title) %>%
  mutate(rel_het = (het - min(het,na.rm=T)) / (max(het,na.rm=T) - min(het,na.rm=T)) ) %>%
  ungroup()
```

Het PCA combo

```{r}
ggplot(inv_het_df) +
  geom_point(aes(x = V1, y = V2, color = het),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free") +
  scale_color_viridis()+
  theme_bw()+
  theme(strip.text = element_text(size=8),
        panel.grid = element_blank())

# ggsave("../../figures/inversions/filter/inversion_pc_plots_heterozygosity.jpeg",
#        width = 20, height = 8)

ggplot(inv_het_df) +
  geom_point(aes(x = V1, y = V2, color = rel_het),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free") +
  scale_color_viridis(name = "Relative Heterozygosity", na.value = "gray40")+
  theme_bw()+
  theme(strip.text = element_text(size=8),
        panel.grid = element_blank())

# ggsave("../../figures/inversions/filter/inversion_pc_plots_relative_heterozygosity.jpeg",
#        width = 20, height = 8)
```

#### Boxplots

This will be very loose - I am going to split PC1 up into 3 for each inversion plot. 
I corrected a few that were obviously not assigned properly by the pc1 three group method.

```{r}
# split pc1 into three parts and assign based solely on where it lands on pc1
inv_karyo <- inv_het_df %>%
  group_by(inversion) %>%
  mutate(karyo = case_when(V1 < (max(V1,na.rm=T) - min(V1,na.rm=T))/3 + min(V1,na.rm=T) ~ "0/0",
                           V1 > 2*(max(V1,na.rm=T) - min(V1,na.rm=T))/3 + min(V1,na.rm=T) ~ "1/1",
                           T ~ "0/1")) %>%
  ungroup()

# check
ggplot(inv_karyo) +
  geom_point(aes(x = V1, y = V2, color = karyo),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free")

# some karyotypes are not properly represented by this...
inv_karyo <- inv_karyo %>%
  mutate(karyo = case_when(strip_title != "chr 1 : 25.1 - 30.7 Mb" ~ karyo,
                           V2 > -0.06 | V1 < 0 ~ "0/1", T ~ "1/1"),
         karyo = case_when(strip_title != "chr 1 : 31.1 - 34.28 Mb" ~ karyo,
                           V2 > 0.3 | V1 < 0.05 ~ "0/0",
                           V1 > 0.25 ~ "1/1", T ~ "0/1"))

# recheck
ggplot(inv_karyo) +
  geom_point(aes(x = V1, y = V2, color = karyo),alpha=0.7) +
  facet_wrap(~ strip_title, nrow=3, scales = "free") 

# ggsave("../../figures/inversions/filter/inversion_pc_plots_karyotypes.jpeg",
#        width = 20, height = 8)
```
#### Significance
Interested in significantly greater heterozygosity in the middle cluster only.
```{r}
library(broom)
library(ggpubr)

# Compare everyone specifically to Pop B
inv_karyo_tests <- inv_karyo %>%
  drop_na(het) %>%
  filter(karyo != "0/1") %>%             # Get the 'others'
  group_by(strip_title, karyo) %>%           # Group by inversion and the 'other' populations
  group_modify(~ {
    # Pull Pop B data for this specific inversion
    het_data <- inv_karyo %>%
      drop_na(het) %>%
      filter(strip_title == .y$strip_title, karyo == "0/1") %>% 
      pull(het)
    
    tidy(wilcox.test(.x$het, het_data, alternative = "less")) # less het in 0/0 & 1/1
  }) %>%
  ungroup() %>%
  mutate(p.adj = p.adjust(p.value, method = "BH"))

# Alternative
labels_df <- inv_karyo_tests %>%
  group_by(strip_title) %>%
  mutate(y_pos = max(inv_karyo$het, na.rm = T) * 1.1) %>%
  ungroup() %>%
  mutate(sig_star = case_when(p.adj < 0.001 ~ "***",
                              p.adj < 0.01  ~ "**",
                              p.adj < 0.05  ~ "*",
                              T ~ "ns"))

ggplot(inv_karyo, aes(x = karyo, y = het, color = karyo)) +
  geom_boxplot() +
  facet_wrap(~ strip_title, nrow=3, scales = "free") +
  geom_text(data = labels_df, 
            aes(x = karyo, y = 0.4, label = sig_star),
            size = 5, vjust = 0, color="black") +
  ylim(c(0,0.45))+
  theme_bw()

# ggsave("../../figures/inversions/filter/inversion_karyotypes_boxplots_signif.jpeg", 
#        width = 20, height = 10)

rm(inv_pca_df, inv_het_df)
```

### Plot by inversion
a) Pop PCA, b) Het PCA, c) Heterozygosity
```{r}
library(patchwork)

for(INV in unique(inv_karyo$strip_title)){

  pca_pop_plot <- ggplot(filter(inv_karyo,strip_title==INV)) +
    geom_point(aes(x = V1, y = V2, color = Pop), alpha=0.7) +
    scale_color_manual(name = "Populations",values=mypalette)+
    theme_bw()+
    labs(x="PC1", y="PC2") +
    theme(strip.text = element_text(size=8),
          panel.grid = element_blank())
  
  pca_het_plot <- ggplot(filter(inv_karyo,strip_title==INV)) +
    geom_point(aes(x = V1, y = V2, color = rel_het),alpha=0.7) +
    scale_color_viridis(name = "Relative\nHeterozygosity", na.value = "gray40")+
    theme_bw()+
    labs(x="PC1", y="PC2") +
    theme(strip.text = element_text(size=8),
          panel.grid = element_blank())
  
  pca_karyo_plot <- ggplot(filter(inv_karyo,strip_title==INV)) +
    geom_point(aes(x = V1, y = V2, color = karyo),alpha=0.7) +
    theme_bw()+
    labs(x="PC1", y="PC2", color = "Inv Type") +
    theme(legend.position="none",
          strip.text = element_text(size=8),
          panel.grid = element_blank())
    
  het_boxplot <- ggplot(filter(inv_karyo,strip_title==INV)) +
    geom_boxplot(aes(x = karyo, y = het, color = karyo)) +
    geom_text(data = filter(labels_df,strip_title==INV), 
              aes(x = karyo, y = 0.4, label = sig_star),
              size = 5, vjust = 0) +
    ylim(c(0,0.45))+
    labs(x = "Putative Inversion Type", y= "Heterozygosity", color = "")+
    theme_bw()
  
  INVNAME=unique((filter(inv_karyo,strip_title==INV))$inversion)
  
  jpeg(str_c("../../figures/inversions/filter/byInv/",INVNAME,".jpeg"), units = "cm", res=300,
       width = 22, height = 20)
  print((pca_pop_plot + pca_karyo_plot) / (pca_het_plot + het_boxplot) + plot_annotation(title=str_c("MDS Outlier Region - ",INV)))
  dev.off()
}
```
