---
title: "boreogadus-selection-scan"
author: "Natasha Howe"
date: "`r Sys.Date()`"
output: html_document
---

PCANGSD Selection Scan

PCAngsd has two implemented selection scans based on (i) FASTPCA and (ii) pcadapt. The associated paper compares the methods with high and low coverage data ([Meisner et al 2021](https://pmc.ncbi.nlm.nih.gov/articles/PMC8480091/#Sec2)).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-metadata, warning=F, message=F}
library(stringr)
library(tools)
library(tidyverse)
library(cowplot)

chr_table <- read_tsv("../../data/R/boreogadus_chromosomes.txt", 
                      col_names = T, show_col_types = F) %>%
  rename(lg = chrName)
```

## Neutral Data

```{r naming}
path= "../../results/pca_selection/"
filename="boreogadus_unfused_neutral_alpha0.05"
outdir="../../figures/pca_selection/"
nEig=4
```

PCA

```{r}
color_df <- read.delim("../../data/R/color_metadata_allpops_flip.txt", header = T)
meta_df <- read.delim("../../data/R/boreogadus_sampleID_popID.txt", sep="\t", 
                      header = F, col.names = c("sampleID","Pop"))
bam_df <- read.table("../../data/bam/boreogadus_filtered_downsampled_bams.txt", header = F)

mypalette <- as.vector(color_df$Color2)
  names(mypalette) <- color_df$Pop # attach pop name to palette color

# Get ID and pop info for each individual
bam_df <- bam_df %>%
  mutate(sampleID = basename(file_path_sans_ext(V1))%>% 
           gsub("boreogadus_|_downsampled|_sorted", "", .)) %>%
  left_join(meta_df, by = "sampleID") %>% 
  mutate(Pop = ifelse(is.na(Pop),"Iceland",Pop)) %>%
  dplyr::select(sampleID, Pop)
  
# read in the covariance matrix
cov <- read.table(paste0(path,filename,"_e",nEig,".cov"))
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)
  e_per <- e$values/sum(e$values) # percent explained by each component

#combine row names (Pop info) with the covariance matrix
pca.vectors = as_tibble(cbind(bam_df, e_vectors))
```

### PCA

```{r}
# manipulate df and plot
pca.vectors[,1:7] %>%
  pivot_longer(cols = starts_with("V"), names_to = "axis", values_to = "val") %>%
  group_by(sampleID) %>% # Replace with your ID column
  mutate(axis_next = lead(axis),
         val_next = lead(val),
         pair = paste(axis, "vs", axis_next)) %>%
  filter(!is.na(val_next)) %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
  ggplot(aes(x = val, y = val_next)) +
  geom_point(mapping=aes(color = Pop),alpha = 0.6) +
  facet_wrap(~pair, scales = "free") +
  scale_color_manual(values = mypalette) +
  labs(x = "Lower PC Axis", y = "Higher PC Axis",
       title=filename) +
  theme_bw()

# ggsave(paste0(outdir,filename,"_PCs1-4.jpg"), height = 6, width = 8)
rm(cov, e, e_per, e_vectors, pca.vectors)
```

### Weights

```{r}
# PC weights
weight_tmp <- read_delim(paste0(path,filename,"_e",nEig,".weights"), delim=" ",
                         col_names = c("pc1","pc2", "pc3","pc4"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(path,filename,"_e",nEig,".sites"), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- read_tsv(str_c(path,filename,".snp_list"), 
                     col_names = c("lg","pos"),
                     show_col_types = F) %>%
  slice(-1) # bc pcangsd removes the first row from the cut beagle file

# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") %>%
  left_join(chr_table, by="lg")

loadings_plot <- weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight), color=as_factor(chr)), size=0.1) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(0, 100, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
  theme_cowplot() +
  labs(x = "Position (Mb)",
       title = "Boreogadus: Neutral Dataset - PCAngsd Loadings") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,filename,"_PC_loadings.jpg"), 
       plot = loadings_plot,
       height = 16, width = 20)

rm(weight_tmp, weight_table)
```

Definitely seems to be inversions driving most of the pc weights form pc1, pc3, and pc4.

### Selection outliers

```{r}
############### nicolas lou rmd
selection_tmp <- read_delim(paste0(path,filename,"_e",nEig,".selection"), delim=" ",
                         col_names = F, show_col_types = F) %>% 
  as_tibble()

# combine selection output with sites
selection_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(selection_tmp) %>%
  pivot_longer(cols = 3:(ncol(selection_tmp)+2), names_to = "pc", values_to = "d") %>%
  mutate(p=pchisq(d, 1, lower.tail=FALSE)) %>%
  left_join(chr_table, by="lg") %>%
  group_by(pc) %>%
  mutate(bonferroni=p.adjust(p, method="bonferroni"),
         fdr=p.adjust(p, method="fdr")) %>%
  ungroup()

# Plot for PC1 and PC2
selection_plot <- selection_table %>%
  ggplot(aes(x=as.numeric(pos)/10^6, y=-log10(p), color=as_factor(chr))) +
  geom_point(size=0.1) +
  geom_point(data=selection_table %>% filter(fdr<0.01), color="red", size=0.2) +
  # geom_point(data=selection_table %>% slice_min(p, n=round(nrow(selection_tmp)/10000)),
  #            color="red", size=0.2) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(10, 90, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
    labs(x = "Position (Mb)",
       title = "Boreogadus: Neutral Dataset Selection Scan") +
  theme_cowplot() +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,filename,"_selection_scan.jpg"), 
       plot = selection_plot, height = 15, width = 20)

rm(selection_tmp)
```

Awesome! PC2 does not have outliers... which that looks like the most neutral trend from before anyways..

### Remove PCA Outlier SNP Regions from Neutral Dataset

It looks like outliers are generally concentrated on one region when found on a chromosome. This means I can remove loci that are within the range of the first and last outlier positions.

```{r}
outlier_regions <- selection_table %>% 
  filter(fdr<0.01) %>%
  group_by(pc,chr) %>%
  summarize(minPos = min(pos), maxPos=max(pos)) %>%
  ungroup() %>%
  arrange(chr, minPos)
  
filtered_sites <- snp_list %>%
  left_join(chr_table, by="lg") %>% 
  left_join(., outlier_regions, by = "chr", relationship = "many-to-many") %>%
  group_by(chr, pos) %>%
  mutate(within_range = any(pos >= minPos & pos <= maxPos, na.rm = TRUE)) %>%
  filter(!within_range) %>%
  ungroup() %>%
  dplyr::select(chr,lg,pos) %>%
  distinct()

print(paste(nrow(filtered_sites),"SNPs left.",
  nrow(snp_list)-nrow(filtered_sites),"SNPs were removed."))

filtered_sites %>%
  mutate(chrpos=str_c(lg,"_",pos)) %>%
  dplyr::select(chrpos) %>%
  write_tsv(str_c("../../data/R/",filename,"_e",nEig,"_rmPCAngsdOutliers.tsv"), col_names=F)
```

### Remove MDS Outlier Regions from Neutral Dataset

Matt shared his MDS outliers output with me. Import here.

```{r}
library(readxl)
library(RColorBrewer)

mds_df <- read_xlsx("../../results/mds/MDS_analysis.xlsx")

offsets <- mds_df %>%
  mutate(CHROM=as.numeric(CHROM)) %>%
  group_by(CHROM) %>%
  dplyr::summarise(max_bp = max(POS, na.rm = T), 
                   .groups = "drop") %>%
  arrange(CHROM) %>%
  mutate(cumulative_bp = cumsum(max_bp),
         offset = lag(cumulative_bp, default = 0),
         chrMid = (max_bp/2 + offset))

# Join back to the main df and calculate cumulative position
mds_df1 <- mds_df %>%
  left_join(offsets, by = "CHROM") %>%
  mutate(cumPos = POS + offset) %>%
  dplyr::select(CHROM,POS,cumPos,starts_with("MD")) %>%
  dplyr::relocate(cumPos, .before="MDS1")

mds_df_longer <- mds_df1 %>%
  pivot_longer(cols = starts_with("MD"),
               names_to = "MDS",
               values_to = "value")

mds_outliers <- mds_df_longer %>%
  group_by(MDS) %>%
  mutate(outlier = value %in% boxplot.stats(value)$out)

out <- boxplot.stats(mds_df1$MDS1)$out
out_ind <- which(mds_df1$MDS1 %in% c(out))
out_ind

outliers <- mds_df1[out_ind,]
View(outliers)

nb.colos <- nrow(offsets)
mycolors <- colorRampPalette(brewer.pal(8,"Set1"))(nb.colos)
names(mycolors) <- offsets$CHROM
```

MDS 1 only

```{r}
ggplot(mds_df1) + 
  # rectangles colored by chrom
  geom_rect(data=offsets, 
            aes(xmin=offset,xmax=cumulative_bp,
                ymin=min(mds_df1$MDS1)-0.01, ymax=max(mds_df1$MDS1)+0.01),
            fill=mycolors,alpha=0.25)+
  # all mds points
  geom_point(data=mds_df1,aes(x=cumPos,y=MDS1,color=as.factor(CHROM)),
             alpha=0.75,cex=0.5)+
  # outlier mds points
  geom_point(data=outliers,aes(x=cumPos,y=MDS1),color="black",cex=0.5)+
  scale_x_continuous(breaks=offsets$chrMid, # offsets$cumulative_bp,
                     labels=offsets$CHROM,
                     expand = c(0.002, 0.002))+
  scale_y_continuous(expand = c(0.001, 0.001))+
  scale_color_manual(values=mycolors)+
  ylab("MDS1")+
  xlab("Chromosome Number")+
  theme_bw() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.text.x = element_text(face="bold",size=6))
```

The outputs looks the same, except that the outlier cutoff is more stringent in his plot.

All MDSs faceted

```{r}
faceted_plot <- ggplot(mds_outliers) + 
  # rectangles colored by chrom
  geom_rect(data = offsets, 
            aes(xmin = offset, xmax = cumulative_bp,
                ymin = -Inf, ymax = Inf, # Use -Inf/Inf to fill the whole height
                fill = as.factor(CHROM)), 
            alpha = 0.25, inherit.aes = FALSE)+
  # all mds points
  geom_point(data=mds_outliers %>% filter(outlier==F),aes(x=cumPos,y=value,color=as.factor(CHROM)),
             alpha=0.75,cex=0.5)+
  # outlier mds points
  geom_point(data=mds_outliers %>% filter(outlier==T),aes(x=cumPos,y=value),color="black",cex=0.5)+
  scale_x_continuous(breaks=offsets$chrMid,
                     labels=offsets$CHROM,
                     expand = c(0.002, 0.002))+
  scale_y_continuous(expand = c(0.001, 0.001))+
  facet_grid(rows=vars(MDS), scales="free_y")+
  scale_color_manual(values=mycolors)+
  scale_fill_manual(values=mycolors)+
  theme_bw() +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.text.x = element_text(face="bold",size=6))

ggsave("../../figures/mds/boreogadus_mds_plots.jpeg",
       plot=faceted_plot, width = 20, height = 25)
```

Didn't end up incorporating the MDS outliers in the new neutral SNP group because I couldn't replicate the cutoff value. Instead, using the regions he identified in the supplemental table as MDS outliers and potential inversions. Included both those that were included and those that weren't. Also removed the identified putative inversions from the Hoff paper.

# Filter Neutral SNPs

Incorporated the MDS outliers Matt tested for inversions & the inversions identified in Hoff Supplemental Table 2 of the pop gen paper.

```{r}
# collected tables that matt created and from Hoff's pop paper
invers_hale <- read_xlsx("../../results/mds/inversion_regions_Hale_and_Hoff.xlsx",sheet = "Hale") %>%
  mutate(chr=as.numeric(chr),startPos = as.numeric(startPos),
         endPos = as.numeric(endPos))
invers_hoff <- read_xlsx("../../results/mds/inversion_regions_Hale_and_Hoff.xlsx",sheet = "Hoff") %>%
  mutate(chr=as.numeric(chr),startPos = as.numeric(startPos),
         endPos =as.numeric(endPos))

inversions <- bind_rows(invers_hale,invers_hoff)

neutral_snps <- filtered_sites %>%
  left_join(., inversions, by = "chr", relationship = "many-to-many") %>%
  group_by(chr, pos) %>%
  mutate(within_inversion = any(pos >= startPos & pos <= endPos, na.rm = TRUE)) %>%
  filter(!within_inversion) %>%
  ungroup() %>%
  distinct(chr,pos)

print(paste(nrow(neutral_snps),"SNPs left.",
  nrow(filtered_sites)-nrow(neutral_snps),"SNPs were removed."))

neutral_snps %>%
  left_join(chr_table, by = "chr") %>%
  mutate(chrpos=str_c(lg,"_",pos)) %>%
  dplyr::select(chrpos) %>%
  write_tsv(str_c("../../data/R/",filename,"_e",nEig,"_rmPCAngsdOutliers_rmInversions.tsv"), col_names=F)

```

### Attempt 2 - Neutral PCA

```{bash eval=F}
sbatch ~/arctic/scripts/bsaida_neutral_rmInversions_pcangsd.sh
```

Using the above snp subset, how does a pca look? Did we get rid of the inversion signal?

```{r}
newfilename="boreogadus_unfused_neutral_alpha0.05_e4_rmPCAngsdOutliers_rmInversions"
nEig=4
#cov <- read.table(paste0("../../results/pca/",newfilename,".cov"))
cov <- read.table(paste0(path,newfilename,"_e",nEig,".cov"))
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)
  e_per <- e$values/sum(e$values) # percent explained by each component
  
#combine row names (Pop info) with the covariance matrix
pca.vectors = as_tibble(cbind(bam_df, e_vectors))# create mypalette

# manipulate df and plot
pca.vectors[,1:7] %>%
  pivot_longer(cols = starts_with("V"), names_to = "axis", values_to = "val") %>%
  group_by(sampleID) %>% # Replace with your ID column
  mutate(axis_next = lead(axis),
         val_next = lead(val),
         pair = paste(axis, "vs", axis_next)) %>%
  filter(!is.na(val_next)) %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
  ggplot(aes(x = val, y = val_next)) +
  geom_point(mapping=aes(color = Pop),alpha = 0.6) +
  facet_wrap(~pair, scales = "free") +
  scale_color_manual(values = mypalette) +
  labs(x = "Lower PC Axis", y = "Higher PC Axis",
       title=newfilename) +
  theme_bw()

ggsave(paste0(outdir,newfilename,"_PCs1-4.jpg"), height = 6, width = 8)
```

#### New Neutral Weights

```{r}
# PC weights
weight_tmp <- read_delim(paste0(path,newfilename,"_e",nEig,".weights"), delim=" ",
                         col_names = c("pc1","pc2", "pc3","pc4"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(path,newfilename,"_e",nEig,".sites"), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- neutral_snps %>%
  left_join(chr_table, by = "chr") %>%
  slice(-1) # bc pcangsd removes the first row from the cut beagle file
  
# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") %>%
  left_join(chr_table, by="lg")

loadings_plot <- weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight), color=as_factor(chr)), size=0.1) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(0, 100, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
  theme_cowplot() +
  labs(x = "Position (Mb)",
       title = "Boreogadus: New Neutral Dataset - PCAngsd Loadings") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,newfilename,"_PC_loadings.jpg"), 
       plot = loadings_plot,
       height = 16, width = 20)

rm(weight_tmp)
```

It looks like the start of chr9 is causing the split for PC1 Let's visualize it closer

```{r}
weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  filter(chr==9, pc=="pc1") %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight)), color="gray20", size=0.8) +
  scale_x_continuous(breaks=seq(0, 100, 1)) 
```

### Attempt 3 - Neutral PCA

Remove snps from chr9 positions 0-6Mb. Rerun pcangsd with that further filtered snp list.

```{bash eval=F}
sbatch ~/arctic/scripts/bsaida_neutral_rmInversions_pluschr9a_pcangsd.sh
```

```{r}
newfilename="boreogadus_unfused_neutral_alpha0.05_e4_rmPCAngsdOutliers_rmInversions_rmchr9a"
nEig=10

#cov <- read.table(paste0(path,newfilename,".cov"))
cov <- read.table(paste0(path,newfilename,"_e",nEig,".cov"))
  e <- eigen(cov)
  e_vectors <- as.data.frame(e$vectors)
  e_per <- e$values/sum(e$values) # percent explained by each component
  varPC1 <- format(round((e$values[1]/sum(e$values)*100), 2), nsmall = 2) 
  varPC2 <- format(round((e$values[2]/sum(e$values)*100), 2), nsmall = 2) 

#combine row names (Pop info) with the covariance matrix
pca.vectors = as_tibble(cbind(bam_df, e_vectors))# create mypalette

# manipulate df and plot
pca.vectors[,1:7] %>%
  pivot_longer(cols = starts_with("V"), names_to = "axis", values_to = "val") %>%
  group_by(sampleID) %>% # Replace with your ID column
  mutate(axis_next = lead(axis),
         val_next = lead(val),
         pair = paste(axis, "vs", axis_next)) %>%
  filter(!is.na(val_next)) %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
  ggplot(aes(x = val, y = val_next)) +
  geom_point(mapping=aes(color = Pop),alpha = 0.6) +
  facet_wrap(~pair, scales = "free") +
  scale_color_manual(values = mypalette) +
  labs(x = "Lower PC Axis", y = "Higher PC Axis",
       title=newfilename) +
  theme_bw()

ggsave(paste0(outdir,newfilename,"_PCs1-4.jpg"), height = 6, width = 8)
```

Yay! Inversions signals are gone. There is still individual pull out, but that's not unexpected. Now only plot PC1

```{r}
pca.vectors %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
  ggplot(aes(x = V1, y = V2)) +
  geom_point(mapping=aes(color = Pop), alpha = 0.6) +
  scale_color_manual(values = mypalette) +
  labs(x = "PC1", y = "PC2") +
  theme_bw() +
  theme(legend.position = "none")

#rm southhampton outliers
pca.vectors %>%
  filter(V1 < 0.5) %>%
  mutate(Pop = factor(Pop, levels = color_df$Pop)) %>%
ggplot(aes(x = V1, y = V2)) +
  geom_point(mapping=aes(color = Pop), alpha = 0.6) +
  scale_color_manual(values = mypalette) +
  labs(x = str_c("PC1 (",varPC1,"%)"), y = str_c("PC2 (",varPC2,"%)"),
       title=newfilename) +
  theme_bw()

```

#### New Neutral Weights

```{r}
# PC weights
weight_tmp <- read_delim(paste0(path,newfilename,"_e",nEig,".weights"), delim=" ",
                         col_names = c("pc1","pc2", "pc3","pc4"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(path,newfilename,"_e",nEig,".sites"), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- snp_list %>%
  mutate(rm = ifelse(chr==9 & pos <=6000000,T,F)) %>%
  filter(rm==F) %>%
  dplyr::select(-rm)
  
# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") %>%
  left_join(chr_table, by="lg")

loadings_plot <- weight_table %>%
  filter(!(str_starts(pc,"X"))) %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight), color=as_factor(chr)), size=0.1) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(0, 100, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
  theme_cowplot() +
  labs(x = "Position (Mb)",
       title = "Boreogadus: New Neutral Dataset - PCAngsd Loadings") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,newfilename,"_PC_loadings.jpg"), 
       plot = loadings_plot,
       height = 16, width = 20)

rm(weight_tmp)
```
