---
title: "matt_code"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = F)
```

```{r install-packages}
library(ggplot2)
library(lostruct)
library(tidyverse)
library(RColoBrewer) 
```

### LD

Plink command for LD

```{bash}
./plink --r2 inter-chr --allow-extra-chr \
  --tfile gls/boreogadus_wholegenome_20241222 \
  --maf 0.05 --ld-window-r2 0.3 \
  -chr OZ177904.1 \
  --out peaks/boreogadus_OZ177904.1_20241222
```

Plot LD

```{r}
redband_OMY05 <- read.delim("/Users/mchale/Desktop/plink/redband_OMY05.ld",sep="",
                            header=T,stringsAsFactors = FALSE)

redband_OMY05 <- redband_OMY05[order(redband_OMY05$R2),]

redband_OMY05 %>% 
  filter(R2>=0.3) %>% 
  ggplot(data=.)+
  geom_point(aes(x=BP_A,y=BP_B,color=R2),shape=15,size=0.7)+
  scale_color_gradient(low="gray99",high="gray4")
```

Natasha edits to ld heatmap plotting

```{r}
CHROM="OZ177908.1"
GROUP <- "Chukchi"
PREFIX <- paste0("boreogadus_",CHROM,"_",GROUP)

matt_ld <- read.delim(paste0("./results/ld/", PREFIX,"_r2.ld"),sep="",header=F,stringsAsFactors = FALSE)
colnames(matt_ld) <- c("BP_A","BP_B","R2")
matt_ld <- matt_ld[order(matt_ld$R2),]

ldplot <- matt_ld %>% 
  filter(R2>=0.3) %>% 
  ggplot(data=.)+
  geom_point(aes(x=BP_A,y=BP_B,color=R2),shape=15,size=0.7)+
  scale_color_gradient()

# Have to select all below and run as one to properly plot in Rmd
tiff(paste0("./figures/ld/", PREFIX,"_LD_r2_plink.tiff"), width = 14, height = 12, res = 300, units = "in", compression = "lzw")
ldplot
dev.off()
```

### MDS - Lostruct

Running MSDS analysis on samples to take into account underlying pop
structure and find regions of the genome that might be behaving
weird/evidence for inversions

1)  Use ANGSD to score SNPs, remember to alter -doGeno to "-doGeno 2" so
    that it spits out genotypes in numeric format i.e., -1, 0, 1, 2.
    this is what's needed as input for MSDS analysis. The most recent
    example is on Charr from Labrador:

```{bash eval=F, echo=T}
./angsd -nthreads 16 -bam labrador2-bam-list.txt \
  -out labrador_charr2 \
  -doMaf 2 -minMaf 0.05 -minInd 100 -doPost 1 -doMajorMinor 1 -snp_pval 1e-6 \
  -doGeno 2 -GL 1 -minQ 20 -minMapQ 20 -doPlink 2 -doCounts 1
```

2)  Run geno output file through Mac's scripts to change the formatting
    of the genotypes and TSV format

```{r}
data_charr2 <- read_tsv("/Users/mchale/Desktop/labrador_charr2.geno.gz",col_names=FALSE)
data_charr2 <- data_charr2[-ncol(data_charr2)]

genos <- data_charr2 %>% 
  select(-X1,-X2) %>% 
  mutate_all(~na_if(.,'-1')) %>% 
  mutate_all(funs(gsub("0","0/0",.))) %>% 
  mutate_all(funs(gsub("1","0/1",.))) %>% 
  mutate_all(funs(gsub("2","1/1",.)))

cols <- data_charr2 %>% 
  select(X1,X2) %>% 
  rename('#CHROM' = X1, POS=X2) %>% 
  mutate(ID="NA") %>% 
  mutate(REF="A",ALT="C",
         QUAL=".",FILTER="PASS",
         "INFO"="NS=181",FORMAT="GT")

vcfbody <- cbind(cols,genos)

write_tsv(vcfbody,"/Users/mchale/Desktop/charr_2.body")
```

Then copy over body file on server and run

```{bash eval=F}
cat redbands.header charr_2.body > charr_2.vcf
bgzip charr_2.vcf
```

and export the zipped vcf back onto the computer (Desktop)

3)  Running MDS

```{r}
snps_charr <- read_vcf("/Users/mchale/Desktop/canada_2.vcf.gz") 
```

NB, you will need `bcftools` installed on the computer and accessible by
R. I have this on my desktop not my laptop because of memory issues...

```{bash}
Sys.setenv(PATH=paste("/Users/matthewhale/Documents/genomics_software/bcftools-1.18/",Sys.getenv("PATH"),sep=":"))
```

Can change the window size depending on number of markers and if RAD seq
or WGS.

```{r}
nWindows = 20
pcs_charr2 <- eigen_windows(snps_charr,k=2,win=nWindows) 
pcdist_charr2 <- pc_dist(pcs_charr2,npc=2)
fit2d_canada2 <- cmdscale(pcdist_charr2,eig=TRUE,k=4)
dim(fit2d_canada2$points)

canada_snps_VCF <- read.vcfR("/Users/mchale/Desktop/canada_2.vcf.gz")
canada_snps_positions <- data.frame(canada_snps_VCF@fix,stringsAsFactors = FALSE)
head(canada_snps_positions)

windows = rep(1:(dim(fit2d_canada2$points)[1]+1),each=nWindows) 
windows <- windows[1:dim(canada_snps_positions)[1]]

canada_snps_positions$windows <- windows
head(canada_snps_positions)
```

```{r}
charr_canada2_chroms <- canada_snps_positions %>% 
  group_by(windows) %>% 
  mutate(Start=min(POS),Stop=max(POS))

write.table(charr_canada2_chroms,"/Users/mchale/Desktop/charr_canada2_chroms.txt",sep="\t")
write.table(fit2d_canada2$points,"/Users/mchale/Desktop/canada2_charr_MDS.txt")
```

Combine window info with MDS and make cumulative bp column in excel.

*NH found error in excel calcs and provides alternative to stay within
R.*

*Error is that the first position was added to the chromosome 1 position
of all subsequent snps when it should just be the basepair position
without any additions for chrom 1. Also, used the second to last
position from chr1 as the counter for chr2, same for chr3.*

```{r}
# Calculate starting offset for each chromosome
offsets <- canada_snps_positions %>%
  mutate(CHROM=as.numeric(CHROM)) %>%
  group_by(CHROM) %>%
  dplyr::summarise(max_bp = max(POS, na.rm = T), 
                   .groups = "drop") %>%
  arrange(CHROM) %>%
  mutate(offset = lag(cumsum(max_bp), default = 0)) %>%
  dplyr::select(CHROM, offset)

# Join back to the main df and calculate cumulative position
charr_canada2_chroms <- canada_snps_positions %>%
  left_join(offsets, by = "CHROM") %>%
  mutate(cumPos = POS + offset) %>%
  dplyr::select(offset) 
```

```{r}
charr_canada2_MDS <- read.delim("/Users/mchale/Desktop/charr_canada2_MDS.txt",sep="\t",header=T,stringsAsFactors = FALSE)
View(charr_canada2_MDS)

# charr_canada2_chroms <- charr_canada2_MDS %>% 
#   group_by(CHROM) %>% 
#   mutate(Start= min(cumulative_bp), 
#          Stop = max(cumulative_bp)) %>% 
#   select(CHROM,Start,Stop) %>% 
#   unique()
```

make Manhatten style plots: NB, Chroms file is redbands_chroms_info.txt

Find the outliers for each MDS:

```{r}
out <- boxplot.stats(charr2_MDS$MDS1)$out
out_ind <- which(charr2_MDS$MDS1 %in% c(out))
out_ind

outliers <- charr2_MDS[out_ind,]
View(outliers)

nb.colos <- nrow (charr_canada2_chroms)
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.colos)
```

good idea to check to make sure number of colours corresponds to the
number of chromosomes.

This is the last step in R to make the Manhattan plot: remember to type
"color=as.factor(Chr))" rather than what Mac has in his protocol or
you'll get a weird error when trying to assign colors to the geom_points

```{r}
ggplot(pinks_MDS_avg) + 
  geom_rect(data=chroms, aes(xmin=Start,xmax=Stop,
                             ymin=min(pinks_MDS_avg$MDS4),
                             ymax=max(pinks_MDS_avg$MDS4)),
            fill=mycolors,alpha=0.25)+
  geom_point(data=pinks_MDS_avg,aes(x=cumulative_bp,y=MDS4,color=as.factor(Chr)),
             alpha=0.75,cex=0.5)+
  geom_point(data=outliers4,aes(x=cumulative_bp,y=MDS4),
             color="black",cex=0.5)+
  scale_x_continuous(breaks=chroms$Stop,labels=chroms$Chr)+
  scale_color_manual(values=mycolors)+
  ylab("MDS4")+
  xlab("Sockeye Linkage group")+
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.text.x = element_text(angle=45,hjust=0.5,face="bold",size=6))

ggsave("/Users/matthewhale/Dropbox/redbands_MDS4_plot.jpg",width=10,height=4)
```

#### Heterozygosity within MDS outliers

Making boxplots to summarize observed heterozygosity for candidate
inversion regions. First, select markers in plink (tped file) that are
in candidate region.

```{bash, eval=F}
./plink --tfile Charr_Canada_LG15_group3 --allow-extra-chr \
  --out canada_charr_LG15_HO \
  --het
```

Annoyingly, PLINK counts the number of homozygotes NOT the number of
hets, so need to do `1-(number of homs - number of SNPs)`. then,
categorize individuals based on genotype. I'm allowing some differences,
but at least 70% of SNPs must show "the same" genotype for me to call
it, in other words, at least 70% of markers must be e.g., homo for the
common form. if that's not the case don't score the individual based on
inversion. then input observed heterozygosity and inversion type into R
and use the obs. heterozygosity to make the boxes using the following
code:

```{r}
my_palette <- colorRampPalette(c("darkblue","blue","lightblue")) (n=3)

ggplot(data=charr_data_boxplot2) + 
  geom_boxplot(mapping=aes(group=inversion.genotype,y = Average.het),fill=my_palette) +
  ylab("average heterozygosity")+ 
  xlab("inversion type")

ggsave("/Users/matthewhale/Documents/lG15_Nunavut_charr_Het_boxplot.jpg",width =5, height = 3)
```

admixture analysis

PCA on candidate CI region with samples colored by population

```{r}
cod_colors <- read.delim("/Users/matthewhale/Dropbox/Arctic_cod/color_metadata_allpops.txt", header=T, sep="\t")

ggplot(data=cod_non_outliers,aes(x=PC2,y=PC3,color=Population))+
  geom_point(alpha=1)+
  scale_color_manual(values=cod_colors$Color)+
  theme(panel.grid = element_blank())
```
